---
title: "Analysis of BLAST Results"
author: "Charlotte Lopp"
date: "October 9, 2017"
output: github_document
---

# Introduction

  It has been established that the diversity of skin-associated bacterial communities has a high enough degree of individual variability that the composition of these communities can be used in forensic indentification. These communities are personaly unique, temporally stable, and transferable. Fierer et al. hypothesized, based on this personalized aspect of skin bacterial communities, that they could use the residual skin bacterial left on objects as "fingerprints" to identify the individual who had touched the object. They proposed to do this by matching the bacteria on the object to the skin-associated bacteria of the individual. In order to perform this matching, three criteria must be met: (1) bacterial DNA recovered from touched surfaces allows for adequate characterization and comparison of bacterial communities; (2) skin bacterial communities persist on surfaces for days to weeks; and (3) surfaces that are touched can be effectively linked to individuals by assessing the degree of similarity between the bacterial communities on the object and the skin of the individual who touched the object. 
  In order to meet these criteria and demonstrate the potential of this approach for forensic identification, three interralted studies were carried out that combined phylogenetic community analyses and high-thorughput pyrosequencing methods. The first experiment entailed comparing bacterial communities on individual keys of three computer keyboards to the communities found on the fingers of the keyboard owners. In the second, the similarity between skin-associated bacterial communities on object stored at -20 degrees Celcius versus those objects stored under typical indoor environmental conditions for up to 14 days were examined. For the third experiment, objects were linked to specific individuals based on comparison of the bacteria on their computer mice to the database containing community information for more than 250 band surfaces, including the hand of the owner. What Fierer et al. discovered was that skin-associated bacteria can be readily recovered from surfaces and that the composition of these communities can indeed be used to determine which individuals touched which objects. This could be determined even if the surfaces had been untouched for up to 2 weeks at room temperature. They demonstrated the utility of a high-throughput pyrosequencing-based approach to quantitatively compare the bacterial communities on objects and skin to match the object to the individual with a high degree of certainty. 

# Methods

## Sample origin and sequencing

Add about a paragraph here.

## Computational

And another paragraph or two here.

# Results

```{r load-libraries, message = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load packages
library("dplyr")
library("tidyr")
library("knitr")
```

```{r make-read-in-data-function}
# Output format from BLAST is as detailed on:
# https://www.ncbi.nlm.nih.gov/books/NBK279675/
# In this case, we used: '10 sscinames std'
# 10 means csv format
# sscinames means unique Subject Scientific Name(s), separated by a ';'
# std means the standard set of result columns, which are:
# 'qseqid sseqid pident length mismatch
# gapopen qstart qend sstart send evalue bitscore',


# this function takes as input a quoted path to a BLAST result file
# and produces as output a dataframe with proper column headers
# and the 'qseqid' column split into sample and seq number
read_blast_output <- function(filename) {
  data_in <- read.csv(filename,
                      header = FALSE, # files don't have column names in them
                      col.names = c("sscinames", # unique Subject Sci Name(s)
                                    "qseqid",    # Query Seq-id
                                    "sseqid",    # Subject Seq-id
                                    "pident",    # Percntge of identical matches
                                    "length",    # Alignment length
                                    "mismatch",  # Number of mismatches
                                    "gapopen",   # Number of gap openings
                                    "qstart",    # Start of alignment in query
                                    "qend",      # End of alignment in query
                                    "sstart",    # Start of alignment in subj
                                    "send",      # End of alignment in subject
                                    "evalue",    # Expect value
                                    "bitscore"))  # Bit score

  # Next we want to split the query sequence ID into
  # Sample and Number components so we can group by sample
  # They originally look like "ERR1942280.1"
  # and we want to split that into two columns: "ERR1942280" and "1"
  # we can use the separate() function from the tidyr library to do this
  # Note that we have to double escape the period for this to work
  # the syntax is
  # separate(column_to_separate,
  # c("New_column_name_1", "New_column_name_2"),
  # "seperator")
  data_in <- data_in %>%
    separate(qseqid, c("sample_name", "sample_number"), "\\.")
}
```

```{r read-in-BLAST-data}
# this makes a vector of all the BLAST output file names, including
# the name(s) of the directories they are in
files_to_read_in <- list.files(path = "output/blast",
                               full.names = TRUE)

# We need to create an empty matrix with the right number of columns
# so that we can rbind() each dataset on to it
joined_blast_data <- matrix(nrow = 0,
                            ncol = 14)

# now we loop over each of the files in the list and append them
# to the bottom of the 'joined_blast_data' object
# we do this with the rbind() function and the function we
# made earlier to read in the files, read_blast_output()
for (filename in files_to_read_in) {
  joined_blast_data <- rbind(joined_blast_data,
                             read_blast_output(filename))
}
```

```{r read-in-metadata-and-join}
# Next we want to read in the metadata file so we can add that in too
# This is not a csv file, so we have to use a slightly different syntax
# here the `sep = "\t"` tells the function that the data are tab-delimited
# and the `stringsAsFactors = FALSE` tells it not to assume that things are
# categorical variables
metadata_in <- read.table(paste0("data/metadata/",
                                 "fierer_forensic_hand_mouse_SraRunTable.txt"),
                          sep = "\t",
                          header = TRUE,
                          stringsAsFactors = FALSE)

# Finally we use the left_join() function from dplyr to merge or 'join' the
# combined data and metadata into one big table, so it's easier to work with
# in R the `by = c("Run_s" = "sample_name")` syntax tells R which columns
# to match up when joining the datasets together
joined_blast_data_metadata <- metadata_in %>%
  left_join(joined_blast_data,
            by = c("Run_s" = "sample_name"))
```


```{r histograms}
# Here we're using the dply piping syntax to select a subset of rows matching a
# criteria we specify (using the filter) function, and then pull out a column
# from the data to make a histogram. We don't need to tell the hist() function
# which data to use, because that's piped in, but we do have to give the
# hist() function the title and axis label we'd like to use for the figure
joined_blast_data_metadata %>%
  filter(env_material_s == "sebum") %>%
  pull(pident) %>%
  hist(main = "Percent Identity",
       xlab = "Percent")
```

Don't forget to report what your figures show in words, here in the Results section.

```{r summary-table}
# Finally, we'd like to be able to make a summary table of the counts of
# sequences for each taxa for each sample. To do that we can use the table()
# function. We add the kable() function as well (from the tidyr package)
# in order to format the table nicely when the document is knitted
kable(table(joined_blast_data_metadata$sscinames,
            joined_blast_data_metadata$Run_s))
```

# Discussion

Add 2-3 paragraphs here interpreting your results and considering future directions one might take in analyzing these data.

