---
title: "Analysis of BLAST Results"
author: "Charlotte Lopp"
date: "October 9, 2017"
output: github_document
---

# Introduction

  It has been established that the diversity of skin-associated bacterial communities has a high enough degree of individual variability that the composition of these communities can be used in forensic indentification. These communities are personaly unique, temporally stable, and transferable. Fierer et al. hypothesized, based on this personalized aspect of skin bacterial communities, that they could use the residual skin bacterial left on objects as "fingerprints" to identify the individual who had touched the object. They proposed to do this by matching the bacteria on the object to the skin-associated bacteria of the individual. In order to perform this matching, three criteria must be met: (1) bacterial DNA recovered from touched surfaces allows for adequate characterization and comparison of bacterial communities; (2) skin bacterial communities persist on surfaces for days to weeks; and (3) surfaces that are touched can be effectively linked to individuals by assessing the degree of similarity between the bacterial communities on the object and the skin of the individual who touched the object.  
  In order to meet these criteria and demonstrate the potential of this approach for forensic identification, three interralted studies were carried out that combined phylogenetic community analyses and high-thorughput pyrosequencing methods. The first experiment entailed comparing bacterial communities on individual keys of three computer keyboards to the communities found on the fingers of the keyboard owners. In the second, the similarity between skin-associated bacterial communities on object stored at -20 degrees Celcius versus those objects stored under typical indoor environmental conditions for up to 14 days were examined. For the third experiment, objects were linked to specific individuals based on comparison of the bacteria on their computer mice to the database containing community information for more than 250 band surfaces, including the hand of the owner. What Fierer et al. discovered was that skin-associated bacteria can be readily recovered from surfaces and that the composition of these communities can indeed be used to determine which individuals touched which objects. This could be determined even if the surfaces had been untouched for up to 2 weeks at room temperature. They demonstrated the utility of a high-throughput pyrosequencing-based approach to quantitatively compare the bacterial communities on objects and skin to match the object to the individual with a high degree of certainty. 

# Methods

## Sample origin and sequencing

  To obtain data for the keyboard study, about 30 individuals keys of three personal computer keyboards and each fingertip of the owner was swabbed. In order to compare the bacterial communities on the keyboards to other miscellaneous keyboards, space bar keys from 15 other private and public computer keyboards were also swabbed. For the storage study, armput skin was simultaneously swabbed with 16 moistened swabs for each of the two individuals. Half of these swabs were immediately frozen at -20 degrees Celsius while the other half was left at room temperature. To gather data for the computer mouse study, the entire exposed face of each computer mouse and the palm surface of nine individual's dominant hands were swabbed. It was important that the mouse had last been touched by the owner 12 hours before swabbing. Fierer et al. estimated the accuracy of matching the house to the owner of the mouse by measuring the degree of similarity between bacterial communities on each computer mouse to the hands of the mouse's owner and to the hands that had never touched the mouse.  
  Genomic DNA was extracted from the swabs using the MO BIO PowerSoil DNA Isolation kit. For each sample, 16S rRNA genes were amplified using the primer set (detailed in "The influence of sex, handedness, and washing on the diversity of hand surface bacteria" by Fierer et al. (2008)) that had been optimized for the phylogenetic analysis of pyrosequencing reads. PCR reactions followed by thermal cycling and aplicon cleaning using UltraClean-htp96-well PCR Clean-up kit (MO BIO) were conducted. Amplicon DNA concentrations were measured using the Quant-it PicoGreen dsDNA reagent and kit (Invitrogen). Pyrosequencing was carried out on a 454 Life Sciences Genome Sequencer FLX instrument (Roche) by the Environmental Genomics Core Facility at the University of South Carolina. 

## Computational

  Computation with this data began with creating a bashscript with code which downloaded the list of 20 files (from the NCBI Sequence Read Archive study number ERP022657) in the run table to the project's data directory. Code was then written to create QC reports for each of the runs using the FastQC program. These 20 sequences were then trimmed based on their quality scores (any sequences below the length of 150 base pairs and any base score below 25 was discarded). The bascript proceeded with code which converted fastq files into fasta files so that they could be used in BLAST queries. Finally, each sequence was blasted with the output written into a csv file; "blastn" was used to search for the top match of each sequence against the *nt* database. These BLAST results are what are now being used for this R-scripted analysis. As the blasting took tens of hours, this session was made sure to be written in a tmux session for the online server. 

# Results

```{r load-libraries, message = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
```

```{r make-read-in-data-function}
# Output format from BLAST is as detailed on:
# https://www.ncbi.nlm.nih.gov/books/NBK279675/
# In this case, we used: '10 sscinames std'
# 10 means csv format
# sscinames means unique Subject Scientific Name(s), separated by a ';'
# std means the standard set of result columns, which are:
# 'qseqid sseqid pident length mismatch
# gapopen qstart qend sstart send evalue bitscore',


# this function takes as input a quoted path to a BLAST result file
# and produces as output a dataframe with proper column headers
# and the 'qseqid' column split into sample and seq number
read_blast_output <- function(filename) {
  data_in <- read.csv(filename,
                      header = FALSE, # files don't have column names in them
                      col.names = c("sscinames", # unique Subject Sci Name(s)
                                    "qseqid",    # Query Seq-id
                                    "sseqid",    # Subject Seq-id
                                    "pident",    # Percntge of identical matches
                                    "length",    # Alignment length
                                    "mismatch",  # Number of mismatches
                                    "gapopen",   # Number of gap openings
                                    "qstart",    # Start of alignment in query
                                    "qend",      # End of alignment in query
                                    "sstart",    # Start of alignment in subj
                                    "send",      # End of alignment in subject
                                    "evalue",    # Expect value
                                    "bitscore"))  # Bit score

  # Next we want to split the query sequence ID into
  # Sample and Number components so we can group by sample
  # They originally look like "ERR1942280.1"
  # and we want to split that into two columns: "ERR1942280" and "1"
  # we can use the separate() function from the tidyr library to do this
  # Note that we have to double escape the period for this to work
  # the syntax is
  # separate(column_to_separate,
  # c("New_column_name_1", "New_column_name_2"),
  # "seperator")
  data_in <- data_in %>%
    separate(qseqid, c("sample_name", "sample_number"), "\\.")
}
```

```{r read-in-BLAST-data}
# this makes a vector of all the BLAST output file names, including
# the name(s) of the directories they are in
files_to_read_in <- list.files(path = "output/blast",
                               full.names = TRUE)

# We need to create an empty matrix with the right number of columns
# so that we can rbind() each dataset on to it
joined_blast_data <- matrix(nrow = 0,
                            ncol = 14)

# now we loop over each of the files in the list and append them
# to the bottom of the 'joined_blast_data' object
# we do this with the rbind() function and the function we
# made earlier to read in the files, read_blast_output()
for (filename in files_to_read_in) {
  joined_blast_data <- rbind(joined_blast_data,
                             read_blast_output(filename))
}
```

```{r read-in-metadata-and-join}
# Next we want to read in the metadata file so we can add that in too
# This is not a csv file, so we have to use a slightly different syntax
# here the `sep = "\t"` tells the function that the data are tab-delimited
# and the `stringsAsFactors = FALSE` tells it not to assume that things are
# categorical variables
metadata_in <- read.table(paste0("data/metadata/",
                                 "fierer_forensic_hand_mouse_SraRunTable.txt"),
                          sep = "\t",
                          header = TRUE,
                          stringsAsFactors = FALSE)

# Finally we use the left_join() function from dplyr to merge or 'join' the
# combined data and metadata into one big table, so it's easier to work with
# in R the `by = c("Run_s" = "sample_name")` syntax tells R which columns
# to match up when joining the datasets together
joined_blast_data_metadata <- metadata_in %>%
  left_join(joined_blast_data,
            by = c("Run_s" = "sample_name"))
```

```{r histograms}
# Here we're using the dply piping syntax to select a subset of rows matching a
# criteria we specify (using the filter) function, and then pull out a column
# from the data to make a histogram. We don't need to tell the hist() function
# which data to use, because that's piped in, but we do have to give the
# hist() function the title and axis label we'd like to use for the figure
library("dplyr")
joined_blast_data_metadata %>%
    filter(env_material_s == "sebum") %>%
    filter(sex_s == "female") %>%
    group_by(sscinames) %>%
    count() %>%
    arrange(desc(n)) %>%
    head(10) %>%
    kable()
```

```{r histogram 2}
library("dplyr")
joined_blast_data_metadata %>%
    filter(env_material_s == "sebum") %>%
    filter(sex_s == "male") %>%
    group_by(sscinames) %>%
    count() %>%
    arrange(desc(n)) %>%
    head(10) %>%
    kable()
      
```

```{r histogram 3}
library("dplyr")
joined_blast_data_metadata %>%
    filter(env_material_s == "sebum") %>%
    filter(sex_s == "male") %>%
    group_by(length) %>%
    ggplot(aes(x = length)) + 
      geom_histogram(color = "gray") +
      theme_classic() +
      ggtitle("Length Histogram for Male and Sebum")
```

This histogram is based on data filtered first by env_material_s "sebum" and then by sex_s "male." What is shown here is a spike in the number of sequences that have a length of just above 150 base pairs. The rest of data is mostly seen in the range between 175-230. 

```{r histogram 4}
joined_blast_data_metadata %>%
    filter(env_material_s == "sebum") %>%
    filter(sex_s == "female") %>%
    group_by(length) %>%
    ggplot(aes(x = length)) + 
      geom_histogram(color = "gray") +
      theme_classic() + 
      ggtitle("Length Histogram for Female and Sebum")
```

This histogram is based on data filtered first by env_material_s "sebum" and then by sex_s "female." What is shown here is a spike in the number of sequences that have a length of just around 225 base pairs. This spike falls within the rest of the data, creating the expected bell curve. Most of the data is within 175-250. 

```{r scatterplot}
ggplot(joined_blast_data_metadata,
 aes(x = length,
 fill = sample_type_s)) +
 geom_density(alpha = 0.5) +
 facet_grid(host_subject_id_s ~ sex_s)
```

This grid of charts shows that in female participants, the greatest density is seen in sequence length of around 225. The strongest spikes are in subjects F5, F6 and F7 but only barely discernable bumps are seen in F2 and F8. In male participants, the greatest density in terms of highest discernable peak is seen in M2 around 225. The only other one that shows the slighest peak at 225 is M9 which also has a peak at just above 150 base pairs. M1, M7 and M8 present peaks at just above 150 base pairs. The computer mouse sample type shows substantially higher peaks at around 225 base pairs in F2, F8 and M2. THe rest of the peaks are only slight. 

```{r intro-to-ggplot}
ggplot(joined_blast_data_metadata,
    aes(x = sex_s,
        y = length)) +
  geom_jitter(aes(alpha = pident,
                  color = sample_type_s)) +
  ggtitle("Pident Scatterplot Across All Sample Types")
```   

This scatterplot is based on pident scores (the percentage of identical matches) for sequence length for all sample types: male, female, not applicable (computer mice). It accounts for pident scores ranging from 85-100%. The strongest scores for females is predominantly between the 200-250 range; the strongest scores for males is just split with a 100 score at just above 150 and then close clustering between about 210-240. The computer mice pident scores reflect the strongest score in males at just above 150 and then the strongest range of scores seen in both sexes between 175 and 250. 

```{r summary-table}
# Finally, we'd like to be able to make a summary table of the counts of
# sequences for each taxa for each sample. To do that we can use the table()
# function. We add the kable() function as well (from the tidyr package)
# in order to format the table nicely when the document is knitted
kable(table(joined_blast_data_metadata$sscinames,
            joined_blast_data_metadata$Run_s))
```

# Discussion

It is interesting to note that the top 10 occurring organisms when compared between male and female palms are considerably different. They in fact only share two: unidentified bacterium and Acridovorax sp. The difference between abundance is also quite distinguishable; for males, Bartonella washoensis’ n is 678 base pairs long while for females, Solemya pervernicosa gill symbiont is 1549. Bartonella washoensis is a bacterium from the genus of Bartonella which was first isolated from a dog with mitral valve endocarditis. It can infect squirrels but can also cause meningitis in humans. According to study Isolation of Bartonella washoensis from a dog with mitral valve endocarditis the main reservoir species for Bartonella washoensis is ground squirrels (Spemophilus beecheyi) from the western United States. Pathologies included progressions from a heart murmur to progressive dyspnea to a diagnosis of congestive heart failure to death a month after the initial presentation. Solemya pervernicosa is one of the 19 species within the genus Solemya, which is a genus of saltwater clams - specifically the awning clams. These clams have chemosynthetic bacterial symbionts that produce their food. These bacteria live within their gill cells and produce energy by oxidizing hydrogen sulfide, which they proceed to use in fixing carbon dioxide via the Calvin cycle. It would be interesting to see why there these, along with some of the other top occurring organisms, appeared on these palms (particularly clams).  
	In the histogram of sequence length for filters of male and sebum, you can see that the highest count is for a length just in just above 150 base pairs. The rest of the count is seen in the range from 175-230. I would guess that this is accounted for by Bartonella washoensis. For the histogram of sequence length for filters of female and sebum, there is a bell curve with the pinnacle being at around the sequence length of 225. The rest of the data is spread around this peak in a range from 175-250. The third grid shows the right palm data for both the female and male sample types and then the computer mouse data for the not applicable (sex) sample type. It is interesting to note the significant peaks seen in F2, F8 and M2 as the magnitude of these peaks are not seen in any of the other mini graphs; the reason as to why these peaks were recorded would be a great avenue to pursue. The scatterplot shows the sequence length but with pident scores (the percentage of identical matches). You can see that for males, the strongest scores are for just above 150 base pairs while for females, the strongest scores are between 200-250. The computer mouse pident scores are strongest at the just above 150 mark and then from about 175-250 accounting for the strength in scores for both sexes. 

